Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-19 paper/main.tex
fill in gaps between theorems

@@

We first consider the estimator \refequ{equ:hutchinson-trace-estimator}. \todo{Introduce $L^1$ norm and Frobenius norm}

@@

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ \todo{introduce spectral norm}. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

@@

\begin{remark}
    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
    \end{equation}
    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
\end{remark}

The Girard-Hutchinson estimator distinguishes itself for its simplicity and provability for all symmetric matrices. However, the $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$. The next two estimators improve on this by exploiting the structure of $\mtx{B}(t)$.

@@

Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank.}

@@

\begin{proof}
    Since $f$ is non-negative, $\mtx{B}(t)$ is positive semi-definite for all $t$. By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$. Therefore, for fixed $t \in [a, b]$
    \begin{equation}
        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
    \end{equation}
    \todo{introduce nuclear norm}.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
    \begin{equation}
        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
    \end{equation}
    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
\end{proof}

@@

\todo{Motivate and explain proof technique (to mimic Hutch++)}

\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right]
        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

@@

    First, we can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
    \end{align}
    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
        \label{equ:matvec-spectral-norm-moment-bound}
    \end{equation}

@@

\begin{proof}
    First, we temporarily omit the parameter dependence. Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e. $\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$. We define the matrix partitions
    \begin{equation}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix}
    \end{equation}
    for some \todo{$k \leq n/2$}. Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.

@@

Finally, we combine the analysis of the two estimators from \refsec{subsec:hutchinson} and \refsec{subsec:nystrom} into a more efficient, general-purpose trace estimator.

@@

\todo{Discuss computation of pseudo inverse: Generalized eigenvalue problem with thresholding \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}, }

>>>>

In the remainder of this section, we will derive upper bounds on the error of these estimators.
%
%\begin{table}[ht]
%\centering
%\renewcommand{\arraystretch}{1.2}
%\begin{tabular}{@{}lcc@{}}
%\toprule
%Estimator & Matrix & $m=1600$\\
%\midrule
%Girard-Hutchinson & symmetric & $\mathcal{O}(\varepsilon^{-2})%$ \\
%Nyström & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-2})$ \\
%Nyström++ & symmetric positive semi-definite & $\mathcal{O}%(\varepsilon^{-1})$  \\
%\bottomrule
%\end{tabular}
%\end{table}

@@

We first consider the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}. We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.

@@

Compared to the constant case \cite[lemma 2.1]{meyer-2021-hutch-optimal} we need to choose the number of queries $n_{\mtx{\Psi}}$ by a factor of $\log(\delta^{-1})$ larger to achieve the same guarantee.

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

@@

\todo{
\begin{remark}
    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
    \end{equation}
    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
\end{remark}
}

The Girard-Hutchinson estimator distinguishes itself for its simplicity. However, its bound does not take into account the structure of $\mtx{B}(t)$. The $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ requirement implies that if we aim to increase the accuracy of the estimate by one digit, we would need to increase the number of queries by a factor of $100$.

@@

For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate. This is reflected in the following theorem.

@@

\begin{proof}
    Since $f$ is non-negative, $\mtx{B}(t)$ is positive semi-definite for all $t$. By \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$. Therefore, for fixed $t \in [a, b]$
    \begin{equation}
        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
    \end{equation}
    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
    \begin{equation}
        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
    \end{equation}
    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.
\end{proof}

\todo{Compare to constant matrix case: \cite[theorem 8.1]{tropp-2023-randomized-algorithms} (similar result)}

The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay, for example $\mtx{B}(t) \equiv \mtx{I}_n$.

@@

Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$ relative error with just $\mathcal{O}(\varepsilon)$ random vectors to construct the estimate. To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[theorem 3.1]{meyer-2021-hutch-optimal}. Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$. Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case. We will do this in a sequence of three lemmas.

For reasons which will become clear in the proof of \reflem{lem:nystrom}, we will need to bound $\mathbb{E}^{p}\left[\lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]$ for a diagonal matrix $\mtx{\Sigma}$ and a standard Gaussian random matrix $\mtx{\Omega}$. Several ways of bounding the moments of $\mtx{\Omega}$ already exist, however, most of them are either not sufficient for our purposes \cite{chen-2005-condition-numbers, edelman-1988-eigenvalues-condition, james-1964-distributions-matrix}. The corresponding result for the Frobenius norm has already been shown in \cite{kressner-2023-randomized-lowrank}. In \cite[lemma B.1]{tropp-2023-randomized-algorithms} a bound is achieved for $p = 2$ and $p = 4$. We will now generalize this result for arbitrary $p \in \mathbb{N}$. We first show the bound for Gaussian random vectors and then extend it to Gaussian random matrices in \reflem{lem:spectral-norm-moment}.

\begin{lemma}{Spectral norm moments of non-standard Gaussian random vector}{spectral-norm-moment-vector}
    The matrix $\mtx{A} \in \mathbb{R}^{m \times m}$ and the standard Gaussian random vector $\vct{\omega} \in \mathbb{R}^{m}$ satisfy for all $k,p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{A} \vct{\omega} \rVert _2 \right]
        \leq  \sqrt{k + p} \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{A} \rVert _F \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

@@

    First, we can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        &\mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] \notag \\
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$, $j=1, \dots, k$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
    \end{align}
    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
        \label{equ:matvec-spectral-norm-moment-bound}
    \end{equation}

@@

\begin{proof}
    First, we temporarily omit the parameter dependence. Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e. $\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$. We define the matrix partitions
    \begin{equation}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix}.
    \end{equation}
    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.

@@

Finally, we can combine the bound on the Girard-Hutchinson estimator \refthm{thm:hutchinson} and the one on the Nyström approximation \reflem{lem:nystrom} to obtain our main result.

@@

The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care.
\todo{Discuss computation of pseudo inverse: Generalized eigenvalue problem with thresholding \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantuma}, }
